When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2018-02-12 paper/paper.tex
Tone down language

\begin{abstract}
    The subgraph isomorphism problem involves finding a copy of a pattern graph inside a larger
    target graph. The non-induced version allows extra edges in the target, whilst the induced
    version does not.  Although both variants are \NP-complete, algorithms inspired by constraint programming
    can operate comfortably on most real-world graphs with thousands of vertices.  However, they
    cannot handle arbitrary instances of this size. We show how to generate ``really hard'' random
    instances for subgraph isomorphism problems, which are computationally challenging with a couple
    of hundred vertices in the target, and only twenty pattern vertices. For the non-induced version
    of the problem, these instances lie on a satisfiable / unsatisfiable phase transition, whose
    location we can predict; for the induced variant, much richer behaviour is observed, and
    constrainedness gives a better measure of difficulty than does proximity to a phase transition.
    These results have practical consequences: we explain why the widely researched ``filter~/
    verify'' indexing technique used in graph databases is a flawed design (based upon a repeated
    misunderstanding of the empirical hardness of \NP-complete problems) that cannot be beneficial
    when paired with any reasonable subgraph isomorphism algorithm.
\end{abstract}

@@

In \cref{section:filterverify} we look at some of the datasets commonly used to test graph database
systems, which the literature suggests are hard to solve. Simple experiments show that the entire
perceived difficulty of each of these datasets comes down to the widespread use of especially poor
subgraph isomorphism algorithms. Our fourth contribution is to show that the simplest constraint
programming approach, even without the more sophisticated recent advances in subgraph algorithms,
makes the entire graph database filter~/ verify paradigm unnecessary.  Finally, we explain why
filter~/ verify and other recently proposed techniques cannot be beneficial even on more challenging
instances except if an extremely poor choice of subgraph isomorphism algorithm is made.

@@

The ``VF2${\leftrightarrow}$'' line shows what happens with VF2 if the graphs are permuted, so that
the non-zero labelled vertices are given lower vertex numbers rather than higher vertex numbers.
In other words, we permute pattern vertices in such a way that the first ten vertices have non-zero
labels, and the ten last vertices have zero labels (and similarly for the target graphs).
This makes VF2 behave better, but still extremely poorly compared to the other two solvers; the
motivation behind this modification will become clear later in this paper.  It will also be
important to remember that both VF2 variations find some satisfiable \emph{and} some unsatisfiable
instances extremely hard.

@@

The filter~/ verify approach has an interesting history, of which we now give a very selective and
incomplete overview. Our description is biased by a general modern understanding of the empirical
hardness of \NP-complete problems, which was not widely known at the time of the earlier
papers we discuss. The common theme of all of the following papers is that pre-computed information
is used to eliminate certain unsatisfiable instances from consideration, without performing a
subgraph isomorphism test. For example, an index might contain a bit of information expressing
whether a target graph contains at least two red vertices. When a pattern graph with two red
vertices is used as a query, any target whose feature set does not have this bit set would not be
considered, and so a subgraph isomorphism call would not be made for that pattern / target pair. (In
practice, feature hashing is often used, which can lead to false positives, but this is not relevant
to our discussion.)

@@

This equation is repeated and expanded upon by \citet{DBLP:journals/tods/YanYH05} to explicitly
separate I/O and isomorphism testing costs, obtaining a query response time of \[
    T_{\mathit{search}} + \left|C_q\right| \cdot (T_{\mathit{io}} + T_{\mathit{iso\_test}})
    \textnormal{.} \] The authors explicitly state that ``the value of $T_{\mathit{iso\_test}}$ does
not change much for a given query''. The argument presented is that
\begin{displayquote}``Sequential scan is very costly
because one has to not only access the whole graph database but also check subgraph isomorphism. It
is known that subgraph isomorphism is an \NP-complete problem. Clearly, it is necessary to build
graph indices in order to help processing graph queries.''\end{displayquote}
With what we now know about the behaviour of modern subgraph isomorphism algorithms, and the nature
of solving \NP-complete problems in general, we should immediately be suspicious of these claims. We
do not expect $T_{\mathit{iso\_test}}$ to be anything like a constant, even if the orders of the
input graphs are similar. In particular, any instance which can be excluded based upon filtering
must have a very small proof of unsatisfiability. These instances \emph{should} be trivial with any
decent subgraph isomorphism algorithm. Thus, all filtering \emph{should} be doing is eliminating the startup
costs of a trivial subgraph isomorphism call. The fact that filtering was successful empirically
should make us wonder whether the subgraph isomorphism algorithm being used was excessively
primitive. Indeed, the subgraph isomorphism algorithm used is described only as ``the simplest approach'' in the
paper.  Additionally, the experiments focus on reducing the size of the candidate set, without
considering the time taken to verify different candidate set instances.  The claim that
$T_{\mathit{iso\_test}}$ does not change much is not justified experimentally, and no consideration
is given as to whether this would hold true for other subgraph isomorphism algorithms.

@@

Rather than the modern techniques used by Glasgow and LAD, we deliberately select a very simple
starting point: a constraint programming model implemented in Gecode 5.1.0 using only
toolkit-provided constraints and heuristics. The model uses a ``smallest domain first'' variable
ordering heuristic with tie-breaking on degree, an all-different constraint, extensional constraints
for adjacency, and labelled-degree filtering at the top of search. Even this simple approach finds
nearly every instance in each of the four datasets trivial, and no exponential behaviour is
observed. As we show in \cref{figure:biiiig-data}, the \emph{hardest} instance for the AIDS dataset
requires 287 recursive calls, and 149,268 of the instances can be solved without search.  For PDBS,
the hardest problem requires 3,161 recursive calls (which occurs 20 times, due to duplicated
queries). For both PCM and PPI, the hardest problem requires 23 recursive calls.  In other words,
none of these instances are in any way computationally hard even for a simple domain-based
algorithm, even before introducing stronger filtering, inference, or heuristics.  Furthermore it is
certainly not the case that a sequential scan is infeasible as so many filter~/ verify papers claim.
In contrast, with VF2, the hardest instance in PDBS requires 31,189 recursive calls; in AIDS,
168,569 calls; in PCMS, 6,476,072 calls; and in PPI, 2,913. In other words, any exponential
behaviour seen in these instances is purely down to a bad choice of subgraph isomorphism algorithm,
and these instances are not inherently ``really hard''.

@@

Unfortunately, this lesson has not been well heeded. For example, \citet{DBLP:journals/tcbb/BonniciG17}
recently compared various non-constraint-based approaches to LAD using benchmark sets whose
difficulties are measured between microseconds and milliseconds. Their results suggest that LAD is
an order of magnitude slower than other approaches. This should not surprise us: the initialisation
costs of LAD are non-trivial due to it constructing domains, calculating neighbourhood degree
sequences, and performing all-different filtering. However, the risks of occasionally encountering
exponential performance from weaker algorithms on easy instances should now be clear.

@@

When labels were introduced, we saw that VF2 finds some instances hard which other algorithms find
easy. We looked at this kind of instance in more detail, and argued that this was due to flaws in
VF2's design, rather than an interesting property of \NP-completeness. Although inevitably there
will be instances that every algorithm finds hard, we see no excuse for an algorithm to exhibit
exponential behaviour in the case when there are two red vertices in a pattern and only one in a
target.

Unfortunately, we saw that this aspect of VF2's performance has had practical consequences, most
notably being the misdesign of graph database systems. By not trying even the simplest constraint
programming techniques from the literature, and ignoring all that is known about the empirical
hardness of \NP-complete problems, members of the graph databases community have reached the mistaken
conclusion that extensive research into supporting techniques is important, whereas really all
they are doing is working around some but not all of the defects in their choices of subgraph
isomorphism algorithms. With this new understanding of what does and does not make subgraph
isomorphism hard, it is time for a radical rethink of how graph database systems work.

We do not claim that the ultimate ``big data'' subgraph matching algorithm already exists:
on the contrary, there is likely to be plenty of room for future improvement now that we understand
the importance of getting algorithm design right.  For example, a major disadvantage of using
domains is the relatively expensive initialisation costs, which quickly add up when dealing with
large numbers of trivial instances. Employing a presolver is an obvious approach---and VF2 is
actually good in this role \citep{DBLP:conf/lion/KotthoffMS16}---but there are other possibilities.
For example, minimal or lazy forward-checking
\citep{DBLP:conf/ictai/DentM94,DBLP:conf/cp/BacchusG95,o:Dent96,DBLP:conf/ecai/LarrosaM98} avoids
constructing every domain upfront, although adopting this require alternatives to ``smallest domain
first'' and to all-different. Such an approach may also be beneficial for huge target graphs, where
having domains range over the entire target is impractical.

>>>>

\begin{abstract}
    The subgraph isomorphism problem involves finding a copy of a pattern graph inside a larger
    target graph. The non-induced version allows extra edges in the target, whilst the induced
    version does not.  Although both variants are \NP-complete, algorithms inspired by constraint programming
    can operate comfortably on most real-world graphs with thousands of vertices.  However, they
    cannot handle arbitrary instances of this size. We show how to generate ``really hard'' random
    instances for subgraph isomorphism problems, which are computationally challenging with a couple
    of hundred vertices in the target, and only twenty pattern vertices. For the non-induced version
    of the problem, these instances lie on a satisfiable / unsatisfiable phase transition, whose
    location we can predict; for the induced variant, much richer behaviour is observed, and
    constrainedness gives a better measure of difficulty than does proximity to a phase transition.
    These results have practical consequences: we explain why the widely researched ``filter~/
    verify'' indexing technique used in graph databases is founded upon a
    misunderstanding of the empirical hardness of \NP-complete problems, and cannot be beneficial
    when paired with any reasonable subgraph isomorphism algorithm.
\end{abstract}

@@

In \cref{section:filterverify} we look at some of the datasets commonly used to test graph database
systems, which the literature suggests are hard to solve. Simple experiments show that the entire
perceived difficulty of each of these datasets comes down to the widespread use of especially poor
subgraph isomorphism algorithms. Our fourth contribution is to show that the simplest constraint
programming approach, even without the more sophisticated recent advances in subgraph algorithms,
makes the entire graph database filter~/ verify paradigm unnecessary.  Finally, we explain why
filter~/ verify and other recently proposed techniques cannot be beneficial even on more challenging
instances, unless the chosen subgraph isomorphism algorithm exhibits unnecessarily poor behavior
on certain classes of instance.

@@

The ``VF2${\leftrightarrow}$'' line shows what happens with VF2 if the graphs are permuted, so that
the non-zero labelled vertices are given lower vertex numbers rather than higher vertex numbers.
In other words, we permute pattern vertices in such a way that the first ten vertices have non-zero
labels, and the ten last vertices have zero labels (and similarly for the target graphs).
This makes VF2 behave better, but still very poorly compared to the other two solvers; the
motivation behind this modification will become clear later in this paper.  It will also be
important to remember that both VF2 variations find some satisfiable \emph{and} some unsatisfiable
instances extremely hard.

@@

The filter~/ verify approach has an interesting history, of which we now give an abridged
overview. Our description is biased by a general modern understanding of the empirical
hardness of \NP-complete problems, which was not widely known at the time of the earlier
papers we discuss. The common theme of all of the following papers is that pre-computed information
is used to eliminate certain unsatisfiable instances from consideration, without performing a
subgraph isomorphism test. For example, an index might contain a bit of information expressing
whether a target graph contains at least two red vertices. When a pattern graph with two red
vertices is used as a query, any target whose feature set does not have this bit set would not be
considered, and so a subgraph isomorphism call would not be made for that pattern / target pair. (In
practice, feature hashing is often used, which can lead to false positives, but this is not relevant
to our discussion.)

@@

This equation is repeated and expanded upon by \citet{DBLP:journals/tods/YanYH05} to explicitly
separate I/O and isomorphism testing costs, obtaining a query response time of \[
    T_{\mathit{search}} + \left|C_q\right| \cdot (T_{\mathit{io}} + T_{\mathit{iso\_test}})
    \textnormal{.} \] The authors explicitly state that ``the value of $T_{\mathit{iso\_test}}$ does
not change much for a given query''. The argument presented is that
\begin{displayquote}``Sequential scan is very costly
because one has to not only access the whole graph database but also check subgraph isomorphism. It
is known that subgraph isomorphism is an \NP-complete problem. Clearly, it is necessary to build
graph indices in order to help processing graph queries.''\end{displayquote}
We reassess these claims in light of
what we now know about the behaviour of modern subgraph isomorphism algorithms and the nature
of solving \NP-complete problems. Contrary to the authors' claims,
we do not expect $T_{\mathit{iso\_test}}$ to be anything like a constant, even if the orders of the
input graphs are similar. In particular, any instance which can be excluded based upon filtering
must have a very small proof of unsatisfiability. These instances \emph{should} be trivial with any
reasonably good subgraph isomorphism algorithm. Thus, all filtering \emph{should} be doing is eliminating the startup
costs of a trivial subgraph isomorphism call. The fact that filtering was successful empirically
should make us wonder whether the subgraph isomorphism algorithm used was substantially
weaker than the state of the art.
Indeed, the algorithm is described only as ``the simplest approach'' in the
paper.  Additionally, the experiments focus on reducing the size of the candidate set, without
considering the time taken to verify different candidate set instances.  The claim that
$T_{\mathit{iso\_test}}$ does not change much is not justified experimentally, and no consideration
is given as to whether this would hold true for other subgraph isomorphism algorithms.

@@

Rather than the modern techniques used by Glasgow and LAD, we deliberately select a very simple
starting point: a constraint programming model implemented in Gecode 5.1.0 using only
toolkit-provided constraints and heuristics. The model uses a ``smallest domain first'' variable
ordering heuristic with tie-breaking on degree, an all-different constraint, extensional constraints
for adjacency, and labelled-degree filtering at the top of search. Even this simple approach finds
nearly every instance in each of the four datasets trivial, and no exponential behaviour is
observed. As we show in \cref{figure:biiiig-data}, the \emph{hardest} instance for the AIDS dataset
requires 287 recursive calls, and 149,268 of the instances can be solved without search.  For PDBS,
the hardest problem requires 3,161 recursive calls (which occurs 20 times, due to duplicated
queries). For both PCM and PPI, the hardest problem requires 23 recursive calls.  In other words,
none of these instances are in any way computationally hard even for a simple domain-based
algorithm, even before introducing stronger filtering, inference, or heuristics.  Furthermore it is
certainly not the case that a sequential scan is infeasible as so many filter~/ verify papers claim.
In contrast, with VF2, the hardest instance in PDBS requires 31,189 recursive calls; in AIDS,
168,569 calls; in PCMS, 6,476,072 calls; and in PPI, 2,913. In other words, any exponential
behaviour seen in these instances is purely down to the choice of subgraph isomorphism algorithm,
and these instances are not inherently ``really hard''.

@@

This lesson has not been well heeded. For example, \citet{DBLP:journals/tcbb/BonniciG17}
recently compared various non-constraint-based approaches to LAD using benchmark sets whose
difficulties are measured between microseconds and milliseconds. Their results suggest that LAD is
an order of magnitude slower than other approaches. This should not surprise us: the initialisation
costs of LAD are non-trivial due to it constructing domains, calculating neighbourhood degree
sequences, and performing all-different filtering. However, the risks of occasionally encountering
exponential performance from weaker algorithms on easy instances should now be clear.

@@

When labels were introduced, we saw that VF2 finds some instances hard which other algorithms find
easy. We looked at this kind of instance in more detail, and argued that this was due to shortcomings in
VF2's design, rather than an interesting property of \NP-completeness. Although inevitably there
will be instances that every algorithm finds hard, we see it as a serious weakness for an algorithm to exhibit
exponential behaviour in such trivial cases as when there are two red vertices in the pattern but only one in the
target.

Unfortunately, we saw that this aspect of VF2's performance has had practical consequences, with the
most notable being the misdesign of graph database systems. By not trying even the simplest constraint
programming techniques from the literature, and disregarding all that is known about the empirical
hardness of \NP-complete problems, members of the graph databases community have reached the mistaken
conclusion that extensive research into supporting techniques is important, whereas really what
they are doing is working around some but not all of the defects in their chosen algorithms for subgraph
isomorphism. With this new understanding of what does and does not make subgraph
isomorphism hard, it is time for a radical rethink of how graph database systems work.

We do not claim that the ultimate ``big data'' subgraph matching algorithm already exists:
on the contrary, there is likely to be plenty of room for future improvement now that we understand
the importance of getting algorithm design right.  For example, a major disadvantage of using
domains is the relatively expensive initialisation costs, which quickly add up when dealing with
large numbers of trivial instances. Employing a presolver is an obvious approach---and VF2 is
in fact good in this role \citep{DBLP:conf/lion/KotthoffMS16}---but there are other possibilities.
For example, minimal or lazy forward-checking
\citep{DBLP:conf/ictai/DentM94,DBLP:conf/cp/BacchusG95,o:Dent96,DBLP:conf/ecai/LarrosaM98} avoids
constructing every domain upfront, although adopting this require alternatives to ``smallest domain
first'' and to all-different. Such an approach may also be beneficial for huge target graphs, where
having domains range over the entire target is impractical.
